CF Registry Analysis - ROC

### 2025-09-15 Christine Chan/CF Registry/cf_registry_ROC_analysis.qmd
EnVision ROC updates for Christine

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
home_dir <- switch(
    Sys.info()["sysname"],
    "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
    "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
    "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
    Sys.info()["sysname"],
    "Darwin" = "/Users/timvigers/GitHub",
    "Windows" = "C:/Users/Tim/GitHub",
    "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

```{r data cleaning}
# Import data
encounter <- read.csv(
    "./Data_Raw/CFF19_encountersMerged_Del1.csv",
    na.strings = c("", "N/A")
)
# Demographic data
demo <- read.csv(
    "./Data_Raw/CFF19_DemogCFDiag_Del1.csv",
    na.strings = c("", "N/A")
)
# Race
races <- c(
    "White",
    "Black or African American",
    "American Indian or Alaska Native",
    "Asian",
    "Native Hawaiian or Other Pacific Islander",
    "Some other race"
)
demo$Race <- apply(demo, 1, function(r) {
    w <- which(r[paste0("Race", 1:6)] == 1)
    if (length(w) == 0) {
        "Unknown"
    } else if (length(w) > 1) {
        "More than one race"
    } else {
        races[w]
    }
})
demo$Race <- factor(
    demo$Race,
    levels = c(
        "White",
        "Black or African American",
        "American Indian or Alaska Native",
        "Asian",
        "More than one race",
        "Native Hawaiian or Other Pacific Islander",
        "Some other race"
    ),
    labels = c(
        "White",
        "Black or African American",
        "Other race",
        "Other race",
        "More than one race",
        "Other race",
        "Other race"
    )
)
demo$Hispanicrace <- factor(
    demo$Hispanicrace,
    levels = 1:2,
    labels = c("Yes", "No")
)
# F508
demo$F508 <- factor(
    demo$F508,
    levels = 1:3,
    labels = c("F508del/F508del", "F508del/Other", "Other/Other")
)
# Age at year end
annual <- read.csv(
    "./Data_Raw/CFF19_Annualized_Del1.csv",
    na.strings = c("", "N/A")
)
annual <- annual %>% rename(reviewyear = ReviewYear)
# Merge
encounter <- left_join(
    encounter,
    demo %>% select(eDWID, Race, Hispanicrace, Gender, F508, MutClass),
    by = join_by(eDWID)
)
encounter <- left_join(
    encounter,
    annual %>% select(eDWID, reviewyear, Age_YrEnd),
    by = join_by(eDWID, reviewyear)
)
# Limit to those with two hour OGTT and HbA1c at the same encounter
# Also remove years before 2010 and those under age 10
roc_df <- encounter %>%
    arrange(eDWID, reviewyear, encounternum) %>%
    select(
        eDWID,
        reviewyear,
        Age_YrEnd,
        Race,
        Hispanicrace,
        Gender,
        F508,
        MutClass,
        twohour_bloodglucose,
        hgba1c
    ) %>%
    filter(
        reviewyear >= 2010,
        Age_YrEnd >= 10,
        hgba1c >= 3.5 & hgba1c <= 15
    ) %>%
    drop_na() %>%
    group_by(eDWID) %>%
    filter(row_number() == n()) %>%
    ungroup() %>%
    mutate(
        `OGTT 2-hour` = cut(
            twohour_bloodglucose,
            breaks = c(-Inf, 140, 200, Inf),
            labels = c("<140", "[140,200)", ">=200")
        ),
        `Age Group` = cut(Age_YrEnd, breaks = c(10, 18, 30, Inf), right = F)
    )
# Metrics to report
metrics <- c(
    "thresh",
    "specificity",
    "sensitivity",
    "accuracy",
    "tn",
    "tp",
    "fn",
    "fp",
    "npv",
    "ppv"
)
```

@@

```{r}
roc_df %>%
    select(
        reviewyear,
        hgba1c,
        Age_YrEnd,
        `Age Group`,
        `OGTT 2-hour`,
        Race,
        Hispanicrace,
        Gender,
        F508,
        MutClass
    ) %>%
    tbl_summary(
        by = `OGTT 2-hour`,
        missing_text = "Missing",
        type = list(reviewyear ~ "categorical"),
        label = list(
            hgba1c = "HbA1c",
            Hispanicrace = "Hispanic",
            MutClass = "Mutation Class",
            reviewyear = "Year",
            Age_YrEnd = "Age at Year End"
        )
    ) %>%
    add_overall() %>%
    add_p(
        test = list(reviewyear = "fisher.test"),
        test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)
    ) %>%
    separate_p_footnotes() %>%
    as_gt()
```

>>>>

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
home_dir <- switch(Sys.info()["sysname"],
    "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
    "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
    "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(Sys.info()["sysname"],
    "Darwin" = "/Users/timvigers/GitHub",
    "Windows" = "C:/Users/Tim/GitHub",
    "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

```{r data cleaning}
setwd(home_dir)
# Import data
encounter <- read.csv(
    "./Data_Raw/CFF19_encountersMerged_Del1.csv",
    na.strings = c("", "N/A")
)
# Demographic data
demo <- read.csv(
    "./Data_Raw/CFF19_DemogCFDiag_Del1.csv",
    na.strings = c("", "N/A")
)
# Race
races <- c(
    "White",
    "Black or African American",
    "American Indian or Alaska Native",
    "Asian",
    "Native Hawaiian or Other Pacific Islander",
    "Some other race"
)
demo$Race <- apply(demo, 1, function(r) {
    w <- which(r[paste0("Race", 1:6)] == 1)
    if (length(w) == 0) {
        "Unknown"
    } else if (length(w) > 1) {
        "More than one race"
    } else {
        races[w]
    }
})
demo$Race <- factor(
    demo$Race,
    levels = c(
        "White",
        "Black or African American",
        "American Indian or Alaska Native",
        "Asian",
        "More than one race",
        "Native Hawaiian or Other Pacific Islander",
        "Some other race"
    ),
    labels = c(
        "White",
        "Black or African American",
        "Other race",
        "Other race",
        "More than one race",
        "Other race",
        "Other race"
    )
)
demo$Hispanicrace <- factor(
    demo$Hispanicrace,
    levels = 1:2,
    labels = c("Yes", "No")
)
# F508
demo$F508 <- factor(
    demo$F508,
    levels = 1:3,
    labels = c("F508del/F508del", "F508del/Other", "Other/Other")
)
# Age at year end
annual <- read.csv(
    "./Data_Raw/CFF19_Annualized_Del1.csv",
    na.strings = c("", "N/A")
)
annual <- annual %>% rename(reviewyear = ReviewYear)
annual <- left_join(annual, demo %>% select(eDWID, Gender), by = join_by(eDWID))
# Get BMI group_by
annual$bmi_cat <- apply(annual, 1, function(r) {
    age <- as.numeric(r["Age_YrEnd"])
    gender <- as.character(r["Gender"])
    bmi <- as.numeric(r["A_bmivalue"])
    bmi_perc <- as.numeric(r["A_bmipercentile"])
    if (age <= 20) {
        return(cut(
            bmi_perc,
            c(-Inf, 10, 50, 85, Inf),
            right = F,
            labels = c("Undernourished", "At Risk", "Target/Goal", "Overweight")
        ))
    } else if (age > 20) {
        if (gender == "M") {
            return(cut(
                bmi,
                c(-Inf, 18.5, 23, 25, Inf),
                right = F,
                labels = c(
                    "Undernourished",
                    "At Risk",
                    "Target/Goal",
                    "Overweight"
                )
            ))
        } else if (gender == "F") {
            return(cut(
                bmi,
                c(-Inf, 18.5, 22, 25, Inf),
                right = F,
                labels = c(
                    "Undernourished",
                    "At Risk",
                    "Target/Goal",
                    "Overweight"
                )
            ))
        }
    }
})
annual$bmi_cat <- relevel(annual$bmi_cat, ref = "Target/Goal")
# Merge
encounter <- left_join(
    encounter,
    demo %>% select(eDWID, Race, Hispanicrace, Gender, F508, MutClass),
    by = join_by(eDWID)
)
encounter <- left_join(
    encounter,
    annual %>%
        select(
            eDWID,
            reviewyear,
            Age_YrEnd,
            A_IsOnEnzymes,
            bmi_cat,
            A_FEV1_pct_predicted,
            A_FVC_pct_predicted
        ),
    by = join_by(eDWID, reviewyear)
)
# Limit to those with two hour OGTT and HbA1c at the same encounter
# Also remove years before 2010 and those under age 10
roc_df <- encounter %>%
    arrange(eDWID, reviewyear, encounternum) %>%
    select(
        eDWID,
        reviewyear,
        Age_YrEnd,
        Race,
        Hispanicrace,
        Gender,
        F508,
        MutClass,
        A_IsOnEnzymes,
        Vx770,
        Vx809comb,
        height,
        weight,
        bmivalue,
        bmipercentile,
        bmi_cat,
        A_FEV1_pct_predicted,
        A_FVC_pct_predicted,
        hgba1c,
        fast_bloodglucose,
        twohour_bloodglucose
    ) %>%
    filter(
        reviewyear >= 2010,
        Age_YrEnd >= 10,
        hgba1c >= 3.5 & hgba1c <= 15
    ) %>%
    group_by(eDWID) %>%
    filter(row_number() == n()) %>%
    ungroup() %>%
    mutate(
        `OGTT 2-hour` = cut(
            twohour_bloodglucose,
            breaks = c(-Inf, 140, 200, Inf),
            labels = c("<140", "[140,200)", ">=200")
        ),
        `Age Group` = cut(Age_YrEnd, breaks = c(10, 18, 30, Inf), right = F)
    )
# Metrics to report
metrics <- c(
    "thresh",
    "specificity",
    "sensitivity",
    "accuracy",
    "tn",
    "tp",
    "fn",
    "fp",
    "npv",
    "ppv"
)
```

@@

```{r}
roc_df %>%
    select(-eDWID) %>%
    tbl_summary(
        by = `OGTT 2-hour`,
        missing_text = "Missing",
        type = list(
            reviewyear ~ "categorical",
            all_continuous() ~ "continuous2"
        ),
        statistic = list(
            all_continuous() ~
                c("{mean} ({sd})", "{median} ({p25}, {p75})", "{min}, {max}")
        ),
        label = list(
            hgba1c = "HbA1c",
            Hispanicrace = "Hispanic",
            MutClass = "Mutation Class",
            reviewyear = "Year",
            Age_YrEnd = "Age at Year End"
        )
    ) %>%
    add_overall() %>%
    add_p(
        test = list(reviewyear = "fisher.test"),
        test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)
    ) %>%
    separate_p_footnotes() %>%
    as_gt()
```
